---
title: "Volume Changed. Lethality Didn't Follow."
subtitle: "NYPD Shooting Incident Data, 2006-present"
author: "Mo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5)
library(tidyverse)
library(lubridate)
library(scales)
```

> Note: `install.packages("tidyverse")` is deliberately **not** in this document.
> Install once from the console; the Rmd only loads libraries so that knitting
> is fast and side-effect free.

# Step 1: Import

**Source:** NYC Open Data, *NYPD Shooting Incident Data (Historic)*,
dataset ID `833y-fsy8`.

**The question this report asks:** New York's shooting volume moved
dramatically over this period. Did the *nature* of those shootings change
alongside the count — and what, if anything, predicts whether a shooting
kills someone?

```{r import}
url_in <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"

nypd_raw <- read_csv(url_in)

dim(nypd_raw)
```

The raw file has `r nrow(nypd_raw)` rows and `r ncol(nypd_raw)` columns.
 
glimpse(nypd_raw)
 
# Step 2: Tidy and Transform

## Summary of the raw data

```{r summary-raw} 
summary(select(nypd_raw, -OCCUR_TIME))
```

```{r na-raw}
colSums(is.na(nypd_raw))
```

## One row is one victim, not one incident

`INCIDENT_KEY` is not unique meaning that when a shooting has more than one
victim, a row (victim) will be created and incident key repeats. As every count 
in this report is a **victim**, reporting these as "incidents" would overstate 
the number of shooting events. Refer to below:

```{r incident-vs-victim}
tibble(
  victim_rows      = nrow(nypd_raw),
  unique_incidents = n_distinct(nypd_raw$INCIDENT_KEY),
  multi_victim_pct = percent(1 - n_distinct(nypd_raw$INCIDENT_KEY) / nrow(nypd_raw),
                             accuracy = 0.1)
)
```


## Missing data and common dirty data caused by data-entry errors

`colSums(is.na())` understates the gaps, because this dataset comes with mixed unknown
values three different ways: `NA`, string `"UNKNOWN"`, and
string `"(null)"`. The age fields contains data-entry
errors, for example `1020`, `224` and `940` that are not reasonable.

```{r inspect-sentinels}
nypd_raw %>% count(PERP_AGE_GROUP, sort = TRUE)
nypd_raw %>% count(VIC_AGE_GROUP,  sort = TRUE)
nypd_raw %>% count(PERP_SEX,       sort = TRUE)
nypd_raw %>% count(LOCATION_DESC,  sort = TRUE) %>% head(10)
```

## The Clean Up

Things that I cleaned up to make the dataset more reasonable and analyzable including 
time of the date without changing the meaning of any rows:

1. Change **`OCCUR_DATE` to a real date**, and a derived `OCCUR_HOUR` so time of day
   can be analysed numerically.
2. **Mixed unknown values input unified to `NA`.** `"UNKNOWN"` and `"(null)"` mean the
   same thing as a blank which should be counted as missing.
3. **Invalid age values dropped to `NA`.** Keep unreasonable 'noise' out.
4. **Geographic coordinate columns removed.** This analysis works at borough
   level, so point coordinates add nothing.
5. **`LOC_OF_OCCUR_DESC` and `LOC_CLASSFCTN_DESC` removed.** These two
   columns are empty for the large majority of the record — see the check below. 
   Keeping them would restrict analysis that used them to recent years only.

```{r late-added-columns}
nypd_raw %>%
  mutate(yr = year(mdy(OCCUR_DATE))) %>%
  group_by(yr) %>%
  summarise(rows = n(),
            missing_loc_class = sum(is.na(LOC_CLASSFCTN_DESC)),
            pct_missing = percent(missing_loc_class / rows)) %>%
  print(n = Inf)
```

```{r tidy}
valid_ages <- c("<18", "18-24", "25-44", "45-64", "65+")

cat_cols <- c("BORO", "PRECINCT", "JURISDICTION_CODE", "LOCATION_DESC",
              "PERP_SEX", "PERP_RACE", "VIC_SEX", "VIC_RACE")

nypd <- nypd_raw %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE),
         OCCUR_HOUR = hour(OCCUR_TIME),
         YEAR       = year(OCCUR_DATE)) %>%
  # mixed unknown values inputs -> NA
  mutate(across(where(is.character),
                ~ if_else(.x %in% c("(null)", "UNKNOWN", "NONE"),
                          NA_character_, .x))) %>%
  mutate(across(c(PERP_SEX, VIC_SEX), ~ na_if(.x, "U"))) %>%
  # invalid age -> NA
  mutate(PERP_AGE_GROUP = factor(if_else(PERP_AGE_GROUP %in% valid_ages,
                                         PERP_AGE_GROUP, NA_character_),
                                 levels = valid_ages),
         VIC_AGE_GROUP  = factor(if_else(VIC_AGE_GROUP %in% valid_ages,
                                         VIC_AGE_GROUP, NA_character_),
                                 levels = valid_ages)) %>%
  mutate(across(any_of(cat_cols), factor)) %>%
  select(-any_of(c("X_COORD_CD", "Y_COORD_CD", "Latitude", "Longitude",
                   "Lon_Lat", "LOC_OF_OCCUR_DESC", "LOC_CLASSFCTN_DESC")))
```

Age brackets are set as **ordered factor levels**.

## Summary of the cleaned-up data

```{r summary-clean}
summary(select(nypd, -OCCUR_TIME))
```

```{r na-clean}
colSums(is.na(nypd))
```

## For the missing data

| Column group | Missing | Treatment |
|---|---|---|
| Victim fields | Very low | Left as-is; negligible effect |
| `JURISDICTION_CODE` | 2 rows | Left as-is |
| `LOCATION_DESC` | High | Not used in the analysis below |
| Perpetrator fields | ~45% | **Rows kept.** Treated as a finding, not a defect — see Bias |

Rows are **not** dropped. Deleting incomplete or dirty records would drop almost
half the data and would increase bias.

# Step 3: Visualize and Analyze

## Finding 1: Volume changed drastically and then boroughs diverged

```{r viz-volume}
by_year_boro <- nypd %>% count(YEAR, BORO)

ggplot(by_year_boro, aes(YEAR, n, color = BORO)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.6) +
  scale_x_continuous(breaks = seq(2006, 2030, 2)) +
  labs(title    = "Shooting victims per year by borough",
       x = NULL, y = "Victims", color = "Borough") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

```{r volume-numbers}
totals <- nypd %>% count(YEAR, name = "victims")

peak_year   <- totals %>% slice_max(victims, n = 1)
trough_year <- totals %>% filter(YEAR < 2020) %>% slice_min(victims, n = 1)
spike_2020  <- totals %>% filter(YEAR %in% c(2019, 2020))

totals %>% print(n = Inf)
```

Shooting volume fell steadily through the 2010s and hit a low of
`r trough_year$victims` victims in `r trough_year$YEAR`, then rose radpidly in
2020 with an increase of
`r percent(spike_2020$victims[2] / spike_2020$victims[1] - 1, accuracy = 1)`
in one year. Every borough rose in 2020. 

After the peak:

```{r recovery}
latest_yr <- max(nypd$YEAR)

recovery <- nypd %>%
  filter(YEAR %in% c(2019, latest_yr)) %>%
  mutate(period = if_else(YEAR == 2019, "pre", "latest")) %>%
  count(BORO, period) %>%
  pivot_wider(names_from = period, values_from = n) %>%
  mutate(change_vs_2019 = percent(latest / pre - 1, accuracy = 1)) %>%
  arrange(desc(latest / pre))
recovery

# years in which the Bronx recorded more victims than Brooklyn
nypd %>%
  filter(BORO %in% c("BRONX", "BROOKLYN")) %>%
  count(YEAR, BORO) %>%
  pivot_wider(names_from = BORO, values_from = n) %>%
  filter(BRONX > BROOKLYN)
```

By `r latest_yr`, some boroughs were back at or below their 2019 level. 
The Bronx recorded more shooting victims than Brooklyn in several recent years 
even with its less dense population.

## Finding 2: Volume and fatality run in opposite directions across the day

```{r viz-hour}
by_hour <- nypd %>%
  group_by(OCCUR_HOUR) %>%
  summarise(victims = n(),
            deaths  = sum(STATISTICAL_MURDER_FLAG),
            rate    = deaths / victims)

by_hour %>%
  select(OCCUR_HOUR, victims, rate) %>%
  pivot_longer(c(victims, rate)) %>%
  mutate(name = recode(name,
                       victims = "Victims (count)",
                       rate    = "Share fatal")) %>%
  ggplot(aes(OCCUR_HOUR, value)) +
  geom_line(linewidth = 0.9, color = "#2F4858") +
  geom_point(size = 1.6, color = "#2F4858") +
  facet_wrap(~name, scales = "free_y", ncol = 1) +
  scale_x_continuous(breaks = seq(0, 23, 3)) +
  labs(title    = "More shootings during the night. More dealy during the day.",
       x = "Hour of day", y = NULL,
       caption  = "Source: NYC Open Data 833y-fsy8") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

```{r hour-numbers}
by_hour %>% mutate(rate = percent(rate, accuracy = 0.1)) %>% print(n = Inf)
```

```{r hour-tests}
hour_cor <- cor.test(by_hour$victims, by_hour$rate)
hour_cor

band_levels <- c("Late night (00-04)", "Morning (05-11)",
                 "Afternoon (12-17)", "Evening (18-23)")

nypd <- nypd %>%
  mutate(TIME_BAND = factor(case_when(
    OCCUR_HOUR <= 4  ~ band_levels[1],
    OCCUR_HOUR <= 11 ~ band_levels[2],
    OCCUR_HOUR <= 17 ~ band_levels[3],
    TRUE             ~ band_levels[4]), levels = band_levels))

by_band <- nypd %>%
  group_by(TIME_BAND) %>%
  summarise(victims = n(),
            deaths  = sum(STATISTICAL_MURDER_FLAG),
            rate    = deaths / victims)
by_band %>% mutate(rate = percent(rate, accuracy = 0.1))

morning_rate <- by_band$rate[by_band$TIME_BAND == "Morning (05-11)"]
rest_rate <- nypd %>%
  filter(TIME_BAND != "Morning (05-11)") %>%
  summarise(r = mean(STATISTICAL_MURDER_FLAG)) %>% pull(r)

morning_test <- with(nypd, chisq.test(table(TIME_BAND == "Morning (05-11)",
                                            STATISTICAL_MURDER_FLAG)))
morning_test
```

Volume peaks late at night: the busiest hour has
`r round(max(by_hour$victims) / min(by_hour$victims), 1)` times as many victims
as the quietest. The fatality rate runs the opposite way. The correlation
between hourly volume and hourly fatality rate is
`r round(hour_cor$estimate, 2)` (p = `r signif(hour_cor$p.value, 2)`).

Aggregating 05:00 to 11:59, morning shootings are fatal
`r percent(morning_rate, accuracy = 0.1)` of the time, against
`r percent(rest_rate, accuracy = 0.1)` for the rest of the day, a ratio of
`r round(morning_rate / rest_rate, 2)`

This observation make me wonder why. Is it fewer bystanders to help, different kinds of
shootings, or traffic during traffic peak hours blocking successful rescue? The model
below tests the last of these.

# Step 4: Bias and Conclusion

**Counts are not rates.** Brooklyn leads every volume chart in this report,
and Brooklyn is also the most populous borough. Without population
denominators these charts partly measure borough size.

```{r caveat-population}
nypd %>%
  count(BORO, sort = TRUE) %>%
  mutate(share = percent(n / sum(n), accuracy = 0.1))
```

Joining NYC population estimates would convert these to per-capita rates and
could reorder the boroughs substantially. That join is the obvious next step
and is deliberately left as an open question rather than papered over.

**The outcome variable is not purely about violence.** Whether a victim dies
depends in part on how quickly they reach ER. `STATISTICAL_MURDER_FLAG`
therefore blends the severity of the shooting with the quality and proximity
of emergency medical care — a confound sitting inside the dependent variable
itself.

**This is police-recorded data.** It captures shootings the NYPD recorded.
Anything not reported to or recorded by the police is absent by construction,
and reporting propensity is unlikely to be uniform across neighbourhoods.



## Conclusion

New York's shooting volume fell by more than half over the 2010s and then
roughly doubled in 2020. Every borough rose together, but they did not come
down together: by `r latest_yr` some had returned to their 2019 levels while
the Bronx and Manhattan remained well above them.

Across the day the relationship is
not just absent but reversed: lethality peaks in the morning, when shootings
are rarest. Victim age is the strongest predictor of a fatal outcome.

The practical reading is that "how many shootings happen" and "how many
shootings kill" are two different questions with two different sets of causes,
and interventions aimed at one should not be expected to move the other.
 
